---
title: "Interactive History Browser"
author: "Lukasz A. Bartnik"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Sub-Processes in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(experiment)
library(knitr)

knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.align = 'center')
knitr::opts_chunk$set(out.width='65%')
```

```{r output_hook,include=FALSE}
(function() {
  hook_output <- knit_hooks$get('output')
  knit_hooks$set(output = function(x, options) {
    if (!is.null(n <- options$out.lines)) {
      if(length(n) == 1) n <- seq(n)
      x  <- unlist(stringr::str_split(x, '\n'))
      lx <- length(x)
      
      dots <- which(diff(n)>1) + 1
      if (length(dots)) {
        x <- x[sort(c(n, dots))]
        x[dots] <- rep('...', length(dots))
      }
      else {
        x <- x[n]
      }
      
      if (min(n) > 1) x <- c('...', x)
      if (max(n) < lx) x <- c(x, '...')
      
      x <- paste(c(x, ""), collapse = '\n')
      options$out.lines <- NULL
    }
    hook_output(x, options)
  })
})()
```

```{r callback,include=FALSE}
knit_hooks$set(evaluate = function(...) {
  res  <- evaluate::evaluate(...)
  plot <- Filter(evaluate::is.recordedplot, res)
  plot <- if (length(plot)) plot[[1]]

  args  <- list(...)
  state <- experiment:::internal_state
  env   <- args$envir
  expr  <- parse(text = paste(args[[1]], collapse = '\n'))[[1]]
  experiment:::update_current_commit(state, env, plot, expr)

  res
})
```

# Introduction

The goal of this tutorial is to introduce the interactive history browser implemented
in the `experiment` package. It follows one of the examples accessible via
`experiment::simulate_london_meters` and is based on the
[London meter data](https://data.london.gov.uk/dataset/smartmeter-energy-use-data-in-london-households).

History browser keeps track of all expressions evaluated in R session.
It remembers all objects and plots, and allows the user to move back and
forth in that recorded history.

In this short introduction, we will perform a simplified data exploration
exercise, similar to what a "real" data exploration might look like. In
order to keep the big picture clean, we avoid poking around too much.

We start by loading a number of packages we will need for our analysis.
History tracker does not write down commands that do not produce new
objects or plots, so it ignores this next block of code.


```{r message=FALSE}
library(dplyr)
library(lubridate)
library(magrittr)
library(ggplot2)
```

# Preparing the data set

Here is the first command that produces a (new) data object. It reads,
transforms and filters a CSV file distributed with the `experiment`
package.

```{r message=FALSE}
input <-
  system.file("extdata/block_62.csv", package = "experiment") %>%
  readr::read_csv(na = 'Null') %>%
  rename(meter = LCLid, timestamp = tstp, usage = `energy_kWh`) %>%
  filter(meter %in% c("MAC004929", "MAC000010", "MAC004391"),
         year(timestamp) == 2013)
```


Let's look at the data. It turns out that the observations are recorded
every 30 minutes.

```{r}
head(input)
```


Let's agregate them and continue with hourly readings.

```{r}
input %<>%
  mutate(timestamp = floor_date(timestamp, 'hours')) %>%
  group_by(meter, timestamp) %>%
  summarise(usage = sum(usage))
```


# The first meter

We have three meters in the data set, `r unique(input$meter)`. We will
look at them one by one, starting with this one.

```{r}
input %<>% filter(meter == "MAC004929")
```


Just a glimpse on the full data set, before we look aggregations.

```{r}
with(input, plot(timestamp, usage, type = 'p', pch = '.'))
```


All right! That doesn't reveal much, how about breaking the data set
down by hour and day of week? Any patterns here? We start with aggregating
the `input` set into a temporary variable `x`.

```{r}
x <-
  input %>%
  mutate(hour = hour(timestamp),
         dow  = wday(timestamp, label = TRUE)) %>%
  mutate_at(vars(hour, dow), funs(as.factor)) %>%
  group_by(hour, dow) %>%
  summarise(usage = mean(usage, na.rm = TRUE))
```


And now we can take a look at the by-hour plot:

```{r}
with(x, plot(hour, usage))
```


And the hour-by-day-of-the-week breakdown:

```{r}
ggplot(x) + geom_point(aes(x = hour, y = usage)) + facet_wrap(~dow)
```


So these are mean values. How about the distribution arund the mean?
We can visualize that with a boxplot. Start with overwriting the `x`
variable and then produce a new plot.

```{r group}
x <-
  input %>%
  mutate(hour = hour(timestamp),
         dow  = wday(timestamp)) %>%
  mutate_at(vars(hour, dow), funs(as.factor))
```

```{r boxplot}
ggplot(x) + geom_boxplot(aes(x = hour, y = usage)) + facet_wrap(~dow)
```


OK! Let's look at a linear model for this data.

```{r}
m <- lm(usage ~ hour:dow, x)
```

```{r summary,out.lines=c(1:14,178:186)}
summary(m)
```


At this point we might decide we know enough. (We probably don't yet, but
for the sake of the presentation, we let's assume we actually do. After
all this is an introduction to the history browser, not to time series
analysis.)


# History recorded so far

So what does the history look like so far? We can open an interactive
viewer by plotting the history object as returned by `fullhistory()`.
It is a `htmlwidget` so when you do it in actual R session in RStudio,
it will open in the `viewer` pane.

```{r fullhistory,fig.width=7,fig.height=5,out.width='100%'}
plot(fullhistory())
```

Each node represents either an object introduced at some point in time
to R session, or a plot. Objects have their names displayed inside the
node, plots are shown as thumbnails.

You can hover your mouse cursor over each node in the history and see
the expression that produced the given object along with its general
characteristics, like dimensions of a `data.frame` or the `AIC` value
for a linear model.


# Second meter

Let's go back to the last step before narrowing down to just one meter.
Clicking on the second `input` node in the history window copies this
object's identifier, which we can then use to bring back the state of
R session just after it was created.


```{r restore}
restore('40692a8b662cc0327e12076193c13a8b730d2fc0')
```


Now we can try a different house.

```{r}
input %<>% filter(meter == "MAC000010")
```

We aggregate the data with the same query as before and look at the
boxplot. Anything interesting here?

```{r group}
```

```{r boxplot}
```

# History, again

The history looks different now, as there is a second branch reflecting
the last three commands we have just issued.

```{r fullhistory,fig.width=7,fig.height=5,out.width='100%'}
```


# Third house

OK, so how about the third house in the data set? We restore the same
point in time again, and repeat the same sequence of commands.

```{r restore}
```

```{r}
input %<>% filter(meter == "MAC004391")
```

```{r group}
```

```{r boxplot}
```

As we can see, the history gets updated again to reflect the third
branching on the third house in the data set.

```{r fullhistory,fig.width=7,fig.height=5,out.width='100%'}
```


# Searching the history

Our last step will be reducing the size of the history graph presented
in the widget. We do it with the `query_by()` function. Let's start with
finding all variables named `input`.


```{r}
h <- query_by(is_named('input'))
```


Looking at the history graph reveals that it is now much smaller.

```{r partial_history,fig.width=7,fig.height=5,out.width='100%'}
plot(h)
```


How about finding only data frames?

```{r}
h <- query_by(inherits('data.frame'))
```

```{r partial_history,fig.width=7,fig.height=5,out.width='100%'}
```


And finally we ask to see only the plots.

```{r}
h <- query_by(inherits('plot'))
```

```{r partial_history,fig.width=7,fig.height=5,out.width='100%'}
```


# Re-building the vignette

In case you have problems when rebuilding this vignette, here is what my
current R session is like:

```{r devtools,message=FALSE}
library(devtools)
devtools::session_info()
```
