---
title: "Supporting Data Processing with an Object Repository"
author: "Lukasz A. Bartnik"
date: "`r Sys.Date()`"
output:
  rmarkdown::pdf_document:
    latex_engine: pdflatex
    pandoc_args: "--latex-engine-opt=--shell-escape"
header-includes:
   - \usepackage{svg}
vignette: >
  %\VignetteIndexEntry{Versioning Data Objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--

Including SVG in PDF generated from RMarkdown:

  https://stackoverflow.com/questions/34064292/is-it-possible-to-include-svg-image-in-pdf-document-rendered-by-rmarkdown


-->


```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(collapse = TRUE, comment = "#>", prompt = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.align = 'center')
knitr::opts_chunk$set(out.width='50%')
```


```{r message=FALSE}
library(readr)
library(lubridate)
library(magrittr)
library(dplyr)
library(ggplot2)
library(broom)
library(modelr)
library(forecast)
library(tidyr)
```


# Dictionary

* `data project`
* `R session`
* `tracker`
* `session cache` (cache)
* `project repository` (repository)
* `artifact`
* `commit`
* `branch`
* `time view`
* `data view`

# Example

## Iterative

Iterate over a single idea, fix errors in the expression, generate a number
of versions of the object and plots, finally choose objects that should make
it to the persistent storage.


### Iterate over a model

Here is a very simple example of what might happen during the initial
phase of a data project. We load the data, build a few simple models
and create a few plots to get the initial insights.


```{r, message=FALSE,warning=FALSE}
file <- 'site_10.csv'
data <- read_csv(file)
```


Let's look at the whole data, one year worth of hourly readings.

```{r, message=FALSE,warning=FALSE}
ggplot(data, aes(x = time, y = usage)) + geom_line()
```


Let's start with monthly averages.

```{r, message=FALSE,warning=FALSE}
m <- data %>%
  mutate(month = as.factor(month(time, label = TRUE))) %>%
  {lm(usage ~ month, .)}
glance(m)
```



There seems to be some regularity in the data.
Let's see if there average usage depends on the day of the week.

```{r, message=FALSE,warning=FALSE}
m <- data %>%
  mutate(wday = wday(time, label = TRUE)) %>%
  {lm(usage ~ wday, .)}
glance(m)
tidy(m)
```


Now let's look at the hourly profile.

```{r, message=FALSE,warning=FALSE}
m <- data %>%
  mutate(hour = as.factor(hour(time))) %>%
  {lm(usage ~ hour, .)}
glance(m)
```

Interesting, both day of the week and the hour of the day can be used
to model the usage. Let's see what happens when we combine them.


```{r, message=FALSE,warning=FALSE}
data %>%
  group_by(hour = hour(time), wday = wday(time, label = TRUE)) %>%
  summarise(usage = mean(usage, na.rm = TRUE)) %>%
  ggplot(aes(x = hour, y = usage)) + geom_point() + facet_wrap(~wday)
```


And now a quantitative assessment of the same model.

```{r, message=FALSE,warning=FALSE}
m <-
  data %>%
  mutate(hour = as.factor(hour(time)), wday = wday(time, label = TRUE)) %>%
  {lm(usage ~ hour:wday, .)}
glance(m)
```

Finally, let's combine all three input features of our model.

```{r, message=FALSE,warning=FALSE}
m <-
  data %>%
  mutate(hour = as.factor(hour(time)), wday = wday(time, label = TRUE),
         month = as.factor(month(time))) %>%
  {lm(usage ~ month + hour:wday, .)}
glance(m)
```



### Browse the history

Up to this point the session cache collected a number of artifacts.
Let's see the summary of the session cache.

```{r eval=FALSE}
tracker
#> Tracker :  ON
#> Branch  :  "default"
#> Commits :  15 in branch, 15 total
```


This is what the history graph looks like. First, in the dimension of
time, that is, where nodes are connected according to the order in
which they were created.

\begin{figure}[h]
\centering
\includesvg[width=140pt]{plot-1}
\caption{svg image}
\end{figure}


Another way to look at the history is the dimension of origin, that is,
the parent-child relationship between artifacts.

\begin{figure}[h]
\centering
\includesvg[width=400pt]{plot-2}
\caption{svg image}
\end{figure}


The information depicted in the first graph can be accessed in R session
via the `tracker$history` facility. The example below follows up on our
sample R session. Commits recorded in the session repository are printed
out starting with the oldest. The most recent commit is assigned index
`a`, the one before is assigned index `b` and so on. Indices can be used
to point to a specific step in the history. For each step we print out
the list of artifact names and the expression abbreviated to one line.

In actual R session variables which are introduced or replaced are printed
in green.


```{r eval=FALSE}
tracker$history(n = 50)
#> o: file
#>    file <- 'site_10.csv'
#> n: data file
#>    data <- read_csv(file)
#> m: data file [plot]
#>    ggplot(data, aes(x = time, y = usage)) + geom_line()
#> l: data file m
#>    m <- data %>% mutate(month = as.factor(month(time, ...
#> k: data file m [printout]
#>    glance(m)
#> j: data file m
#>    m <- data %>% mutate(wday = wday(time, label = TRUE)) ...
#> i: data file m [printout]
#>    glance(m)
#> h: data file m [printout]
#>    tidy(m)
#> g: data file m
#>    m <- data %>% mutate(hour = as.factor(hour(time))) ...
#> f: data file m [printout]
#>    glance(m)
#> e: data file m [plot]
#>    data %>% group_by(hour = hour(time), wday = wday(time, ...
#> d: data file m
#>    m <- data %>% mutate(hour = as.factor(hour(time)), ...
#> c: data file m [printout]
#>    glance(m)
#> b: data file m
#>    m <- data %>% mutate(hour = as.factor(hour(time)), ...
#> a: data file m [printout]
#>    glance(m)
```


Each commit can be printed in more detail. This is where the letters-indices
come in handy.

```{r eval=FALSE}
tracker$history$b
#> Commit afb67ad6
#> data :  data.frame[8784, 2]
#> file :  character(1)
#> m    :  lm(formula = usage ~ month + hour:wday, data = .)
#>
#>   m <-
#>    data %>%
#>    mutate(hour = as.factor(hour(time)), wday = wday(time, label = TRUE),
#>           month = as.factor(month(time))) %>%
#>    {lm(usage ~ month + hour:wday, .)}
```


Since we have found an interesting artifact, we can now instruct `tracker`
to store it in the persistent repository of artifacts. We point to one
object but the whole sequence of objects required to create it is copied
from the session repository to the persistent one. If any of the objects
is already present in the persistent repository (`session cache` vs `repository`)
the ends are connected and a consistent graph of artifacts is maintained.



```{r eval=FALSE}
tracker$history$b$m %>% persist
#> Persisted 3 objects: m, data, file
```


Let's all persist the final profile plot.

```{r eval=FALSE}
tracker$history$e$`[plot]` %>% persist
#> Persisted 1 plot.
```



## Repeat the Sequence

Another useful tool is the `replay` function. It accepts a list of objects
that were changed, finds their originals, and re-computes all nodes descendant
in the `data` view. In our example we replace the data file name and then
ask R to recalculate all object below it. It creates a new branch in the
session cache and names it `"replay(file) #1"`.


```{r eval=FALSE}
file <- 'site_88.csv'
replay(file)
#> Objects changed: file
#> Recalculating: 6 objects, 2 plots, and 6 printouts.
#> Branch name: "replay(file) #1"
```


Let's look at the state of the tracker.

```{r eval=FALSE}
tracker
#> Tracker :  ON
#> Branch  :  "replay(file) #1"
#> Commits :  15 in branch, 30 total
```


The current R session contains a new model, built with the same pipeline
but for a different data set. Let's take a look at the model and compare
it to the one already stored in the project repository. The new model:

```{r eval=FALSE}
glance(m)
```
```{r echo=FALSE}
file <- 'site_88.csv'
data <- read_csv(file)
m2 <-
  data %>%
  mutate(hour = as.factor(hour(time)), wday = wday(time, label = TRUE),
         month = as.factor(month(time))) %>%
  {lm(usage ~ month + hour:wday, .)}
glance(m2)
```

The model built for __site 10__:

```{r eval=FALSE}
glance(tracker$repository$m)
```
```{r echo=FALSE}
glance(m)
```

Let's also take a look at the final profile plot for __site 88__. In the
session history it can be accessed via:

```{r eval=FALSE}
tracker$history$e$`[plot]`
```
```{r echo=FALSE,message=FALSE,warning=FALSE}
data %>%
  group_by(hour = hour(time), wday = wday(time, label = TRUE)) %>%
  summarise(usage = mean(usage, na.rm = TRUE)) %>%
  ggplot(aes(x = hour, y = usage)) + geom_point() + facet_wrap(~wday)
```

Look back, change input, generate two new models for two other buildings;
use `repeat`


Browse new branches, choose the two new models and store them persistently


## Browse the Tree of Understanding

Look at the branches stored in the persistent store, this time presented
according to their origin



